### 2015-02-04 CICM2015/lemmaclustering.tex
Renamed

ML4PG uses (unsupervised) \emph{clustering algorithms}~\cite{Bishop} to find patterns in Coq syntax and proofs.
Clustering algorithms % is a set of machine-learning techniques that 
divide data into $n$ groups of similar objects (called \emph{clusters}), where the value of $n$
is a parameter provided by the user. In ML4PG, the value of $n$ is automatically computed depending on the number 
of objects to cluster, and using the formula provided in~\cite{KHG13,lpar13}. 
A detailed exposition of machine-learning algorithms involved in ML4PG can be found in \cite{KHG13}. In this paper, our first goal consists in improving ML4PG's feature extraction algorithm.
%In this section, we present a method to collect features from Coq terms.

\emph{Feature extraction}~\cite{Bishop} is a research area developing methods for 
discovery of statistically significant features in data. % in machine-learning, known as . 
We adopt the following standard terminology. % concerning feature extraction. 
We assume there is a training data set, containing some samples (or objects). 
\emph{Features} are given by a set of statistical parameters chosen to represent all objects in the given data set.
If $n$ features are chosen, one says that object classification is 
conducted in an $n$-dimensional space. 
For this reason, most pattern-recognition tools will require that %the features have numeric values; and 
the number of selected features is limited and fixed (sparse methods, like the ones applied in
e.g.~\cite{lpar-urban,K13,UrbanSPV08} are the exception to this rule). 
\emph{Feature values} are rational numbers used to instantiate the features for every given object.  
If an object is characterised by $n$ feature values, these $n$ values together form a \emph{feature vector} for this object. 
%Irrespective of the particular algorithm
%used, 
A function that assigns, to every object of the data set, a feature vector is called a \emph{feature extraction function}.
Normally, feature extraction is a data pre-processing stage, and it is separated from the actual pattern-recognition process. %ML4PG relies on unsupervised clustering methods~\cite{KHG13}.

Feature extraction from terms or \emph{term trees} is common to most feature extraction algorithms implemented
in theorem provers~\cite{lpar13,lpar-urban,K13,UrbanSPV08}. In~\cite{lpar13}, we introduced a feature extraction mechanism
for ACL2 first-order terms. Here, that ACL2 method is substantially re-defined to capture the higher-order dependently-typed language of Coq.
%However, due to the different nature of ACL2 and Coq -- ACL2 being a first-order untyped language, and Coq being a higher-order
%dependently-typed language -- the algorithm presented in~\cite{lpar13} cannot be extrapolated to Coq, and a different method is necessary.
%Therefore, we introduce here a new feature-extraction method for Coq terms.

The underlying formal language of Coq is known as the \emph{Predicative Calculus of (Co)Inductive Constructions} (pCIC)~\cite{Coq,CoquandH88,CoPa89}. % -- a 
%weaker calculus of the \emph{Calculus of (Co)Inductive Constructions}. 
The terms of pCIC are built from
the following rules:

@@

The syntax of Coq terms~\cite{Coq} includes some terms that do not appear in Definition~\ref{def:coqterms}; e.g. given 
 a variable \lstinline?x?, and  terms \lstinline?T? and \lstinline?U?, \lstinline?fix name (x:T) := U? is a Coq term used to declare a recursive definition.
The notion of a term in Coq covers a very general syntactic
category in the Gallina specification language~\cite{Coq} and corresponds to the intuitive notion of well-formed expression. 
%Our feature extraction
%method works for all well-formed expressions that can be defined in Coq; 
However, for the purpose of concise exposition, % purposes,
we will restrict our notion of a term to Definition~\ref{def:coqterms},
giving the full treatment of the whole Coq syntax in the actual ML4PG implementation.
 % %definitions to the syntax of Coq terms coming from pCIC terms, thereby remembering that full Coq syntax is handled in the actual ML4PG implementation.

@@

Note that ML4PG term trees consist of two kinds of nodes: \emph{Gallina} and \emph{term-type} nodes. The Gallina nodes are 
labelled by Gallina-keywords and special tokens such as \lstinline?forall?, \lstinline?fun?, \lstinline?let? or \lstinline?->? (from now on, we will call them Gallina tokens); 
%to
%Gallina keywords and special tokens); 
and the term-type nodes are labelled by expressions of the form \lstinline?t1:t2? where \lstinline?t1? is a sort, a variable or 
a name, and  \lstinline?t2? is the type of \lstinline?t1?.

@@

We represent ML4PG term trees as feature matrices, further to be flattened into feature vectors for clustering. %they in turn will be the input of machine-learning algorithms.
A variety of methods exists to represent trees as matrices, for instance using
adjacency or incidence matrices. The adjacency matrix and the various previous methods of  feature
extraction (e.g.~\cite{lpar-urban,K13,UrbanSPV08}) share the following common properties: different %tree
%nodes 
library symbols are represented by distinct features, and the %matrix entries
feature values are binary. %; and the size of the matrix depends on the tree size. 
For
big libraries and growing syntax, feature vectors grow very large (up to $10^6$ in some experiments)
and at the same time very sparse, which implies the use of sparse machine-learning methods.

%In~\cite{lpar13}, we represent ACL2 term trees as matrices with the following properties: the features are 
%given by the arity of functions and the term tree depth -- two properties common to every term tree -- and the 
%feature values are rational numbers dynamically computed depending on the definition of the functions involved in the term tree. The size of 
%those matrices is fixed to $49$; hence, allowing the application of data-mining techniques to both big and small 
%data sets of ACL2 theorems -- large and sparse feature vectors require to use big data sets, 
%where the size of the example data set is comparable to the number of features; on the contrary, compact feature extraction methods work equally well 
%with both big and small datasets. The method presented in~\cite{lpar13}
%can be applied not only to ACL2 but also to any first-order theory; however, it is not sufficient to capture all the information 
%from the higher-order dependently-typed language of Coq.

We develop a new compact method that tracks a large (potentially unlimited) number of Coq terms by a finite number of features and an unlimited number of feature-values.
In our method, the features are given by two properties common to all possible term trees: the term tree depth and the level index of nodes.
The most important information about the term is then encoded by improving precision of feature values %means of a more precise specification of feature values, 
using rational-valued feature-extraction functions. Taking just $300$ features, the new feature-extraction method recursively adjusts the feature values, 
adapting to the growing language syntax. %; thus producing a machine-learning analogue of how an unlimited number of Coq terms can be generated from the limited syntactic constructs.  
The resulting feature vectors have an average density ratio of 60\%.

@@

	%%% Introduce common notation for all three; may be call it more like feature exatraction... template f.e.? pre-fe.e?

% \begin{definition}[ML4PG term tree matrix]\label{df:matrix}
% Given a term tree $T$ for a Coq term, and three injective functions $[.]_{Gallina}: Gallina~tokens \rightarrow \mathbb{Q}^-$, $[.]_{term}: Coq~Terms \rightarrow \mathbb{Q}^+$, $[.]_{type}: Coq~Terms \rightarrow \mathbb{Q}^+$, 
% the ML4PG term tree matrix $M_M$ of $T$ is a $10 \times 10$ matrix such that the $(i,j)$-th entry of $M_M$ captures information from the node of depth $i$ and
% level index $j$ of $T$ as follows:
% 
% $-$ if the node of depth $i$ and level index $j$ of $T$ is a Gallina node $g$, then the $(i,j)$-th entry of $M_M$ is a triple $([g]_{Gallina},-1,p)$ 
% where $p$ is the level index of the parent of $g$. %%% --- if $c$ is the root node, the value of $p$ is $-1$. 
% 
% $-$ if the node of depth $i$ and level index $j$ of $T$ is a term-type node \lstinline?t1:t2?, then the $(i,j)$-th entry of $M_M$ is a triple $([t1]_{term},[t2]_{type},p)$
% where $p$ is the level index of the direct parent of the node.
% 
% \end{definition}

@@

$-$ if $T_{\texttt{t}}(i,j)$ is a Gallina node $g$, then the $(i,j)$th entry of $[\texttt{t}]_M$ is a triple $([g]_{Gallina},-1,p)$ 
where $p$ is the level index of the parent of $g$. %%% --- if $c$ is the root node, the value of $p$ is $-1$.

@@

In the above definition, we fix  the maximum depth and  maximum level index of a node to $10$; this makes the feature extraction mechanism uniform
across all Coq terms appearing in the libraries. We may lose some information if  pruning is needed, but the chosen size works well for most terms appearing in Coq libraries. 
%The above method works for any Coq term, irrespective of the size and complexity. 
If a term tree does not fit into $10 \times 10$ term tree dimensions, its $10 \times 10$
subtree is still considered by ML4PG.
%For the machine-learning algorithms, 
The term tree matrix is flattened into a feature vector, and each triple will be split 
into three components of the vector, giving a feature vector size of $300$, %-- bigger than the size of vectors in~\cite{lpar13}, 
still smaller than in sparse approaches~\cite{lpar-urban,K13,UrbanSPV08}.

In Definition~\ref{df:matrix}, we deliberately specify the functions $[.]_{Gallina}, [.]_{term}$ and $[.]_{type}$ just by their signature. 
The function $[.]_{Gallina}$ is a predefined function.
The number of Gallina tokens (\lstinline?forall?, \lstinline?fun?, \lstinline?->? and so on) is fixed and cannot be 
expanded by the Coq user. Therefore, we know in advance all the Gallina tokens that can appear in a development, and we can
assign a concrete value to each of them. The function $[.]_{Gallina}: Gallina~tokens  \rightarrow \mathbb{Q}^-$ is an injective 
function carefully defined to assign close values to similar Gallina tokens %(for instance, 
%$[$\lstinline?forall?$]_{Gallina}=-1$ and $[$\lstinline?->?$]_{Gallina}=-1.1$);
and more distant numbers to unrelated tokens 
%(e.g. $[$\lstinline?let?$]_{Gallina}=-3$)
--- see Appendix~\ref{sec:gallinasyntax} for the exact encoding.
 %; thus distinguishing the 
%different kinds of Gallina tokens unambiguously.

%On the contrary to the function $[.]_{Gallina}$, 
The functions $[.]_{term}$ and $[.]_{type}$ are dynamically re-defined for every 
library and every given proof stage, to adapt to the changing syntax.
 %, as we are going to describe in 
%Definition~\ref{def:funterm}. 
In practice, there will be  %set of functions computed in 
%every session: 
new $[.]_{term}$ and $[.]_{type}$ functions computed whenever ML4PG is called.
This
brings the element of the ``acquired knowledge/experience'' to the machine-learning cycle, as will be formalised in the next section.

% In the previous section, we have presented a method to extract features from 
% Coq terms; however, the functions $[.]_{Gallina}$, $[.]_{term}$ and $[.]_{type}$
% are still undefined. 
% % The definition of these functions influences the results that 
% % are obtained by machine-learning algorithms.
% In this section, we explain how these functions are defined and how this has led to several improvements in ML4PG.

%We finish this section with the following property about ML4PG term tree matrices.

% \begin{lemma}
% Let ${\cal T}$ be the set of Coq terms whose nodes have maximum depth $10$ and level index $10$. 
% Then, the function $[.]_{T}$ restricted to ${\cal T}$ is a one-to-one function. 
% %  Given a term tree matrix $M$ associated with a Coq term $C$. If the maximum depth of the term tree of $C$ is 10 and the maximum index level of the 
% %  nodes of the tree is $10$, then the term $C$ can be reconstructed from $M$. 
% \end{lemma}

>>>>

ML4PG uses (unsupervised) \emph{clustering algorithms}~\cite{Bishop} to find patterns in Coq syntax and proofs.
Clustering algorithms
divide data into $n$ groups of similar objects (called \emph{clusters}), where the value of $n$
is a parameter provided by the user. In ML4PG, the value of $n$ is automatically computed depending on the number
of objects to cluster, and using the formula provided in~\cite{KHG13,lpar13}.
A detailed exposition of machine-learning algorithms involved in ML4PG can be found in \cite{KHG13}. In this paper, our first goal consists in improving ML4PG's feature extraction algorithm.

\emph{Feature extraction}~\cite{Bishop} is a research area developing methods for
discovery of statistically significant features in data.
We adopt the following standard terminology.
We assume there is a training data set, containing some samples (or objects).
\emph{Features} are given by a set of statistical parameters chosen to represent all objects in the given data set.
If $n$ features are chosen, one says that object classification is
conducted in an $n$-dimensional space.
For this reason, most pattern-recognition tools will require that
the number of selected features is limited and fixed (sparse methods, like the ones applied in
e.g.~\cite{lpar-urban,K13,UrbanSPV08} are the exception to this rule).
\emph{Feature values} are rational numbers used to instantiate the features for every given object.
If an object is characterised by $n$ feature values, these $n$ values together form a \emph{feature vector} for this object.
A function that assigns, to every object of the data set, a feature vector is called a \emph{feature extraction function}.
Normally, feature extraction is a data pre-processing stage, and it is separated from the actual pattern-recognition process.

Feature extraction from terms or \emph{term trees} is common to most feature extraction algorithms implemented
in theorem provers~\cite{lpar13,lpar-urban,K13,UrbanSPV08}. In~\cite{lpar13}, we introduced a feature extraction mechanism
for ACL2 first-order terms. Here, that ACL2 method is substantially re-defined to capture the higher-order dependently-typed language of Coq.

The underlying formal language of Coq is known as the \emph{Predicative Calculus of (Co)Inductive Constructions} (pCIC)~\cite{Coq,CoquandH88,CoPa89}.
The terms of pCIC are built from
the following rules:

@@

The syntax of Coq terms~\cite{Coq} includes some terms that do not appear in Definition~\ref{def:coqterms}; e.g. given
 a variable \lstinline?x?, and  terms \lstinline?T? and \lstinline?U?, \lstinline?fix name (x:T) := U? is a Coq term used to declare a recursive definition.
The notion of a term in Coq covers a very general syntactic
category in the Gallina specification language~\cite{Coq} and corresponds to the intuitive notion of well-formed expression.
However, for the purpose of concise exposition,
we will restrict our notion of a term to Definition~\ref{def:coqterms},
giving the full treatment of the whole Coq syntax in the actual ML4PG implementation.

@@

Note that ML4PG term trees consist of two kinds of nodes: \emph{Gallina} and \emph{term-type} nodes. The Gallina nodes are
labelled by Gallina-keywords and special tokens such as \lstinline?forall?, \lstinline?fun?, \lstinline?let? or \lstinline?->? (from now on, we will call them Gallina tokens);
and the term-type nodes are labelled by expressions of the form \lstinline?t1:t2? where \lstinline?t1? is a sort, a variable or
a name, and  \lstinline?t2? is the type of \lstinline?t1?.

@@

We represent ML4PG term trees as feature matrices, further to be flattened into feature vectors for clustering.
A variety of methods exists to represent trees as matrices, for instance using
adjacency or incidence matrices. The adjacency matrix and the various previous methods of  feature
extraction (e.g.~\cite{lpar-urban,K13,UrbanSPV08}) share the following common properties: different
library symbols are represented by distinct features, and the
feature values are binary.
For
big libraries and growing syntax, feature vectors grow very large (up to $10^6$ in some experiments)
and at the same time very sparse, which implies the use of sparse machine-learning methods.

We develop a new compact method that tracks a large (potentially unlimited) number of Coq terms by a finite number of features and an unlimited number of feature-values.
In our method, the features are given by two properties common to all possible term trees: the term tree depth and the level index of nodes.
The most important information about the term is then encoded by improving precision of feature values
using rational-valued feature-extraction functions. Taking just $300$ features, the new feature-extraction method recursively adjusts the feature values,
adapting to the growing language syntax.
The resulting feature vectors have an average density ratio of 60\%.

@@

@@

$-$ if $T_{\texttt{t}}(i,j)$ is a Gallina node $g$, then the $(i,j)$th entry of $[\texttt{t}]_M$ is a triple $([g]_{Gallina},-1,p)$
where $p$ is the level index of the parent of $g$.

@@

In the above definition, we fix  the maximum depth and  maximum level index of a node to $10$; this makes the feature extraction mechanism uniform
across all Coq terms appearing in the libraries. We may lose some information if  pruning is needed, but the chosen size works well for most terms appearing in Coq libraries.
If a term tree does not fit into $10 \times 10$ term tree dimensions, its $10 \times 10$
subtree is still considered by ML4PG.
The term tree matrix is flattened into a feature vector, and each triple will be split
into three components of the vector, giving a feature vector size of $300$,
still smaller than in sparse approaches~\cite{lpar-urban,K13,UrbanSPV08}.

In Definition~\ref{df:matrix}, we deliberately specify the functions $[.]_{Gallina}, [.]_{term}$ and $[.]_{type}$ just by their signature.
The function $[.]_{Gallina}$ is a predefined function.
The number of Gallina tokens (\lstinline?forall?, \lstinline?fun?, \lstinline?->? and so on) is fixed and cannot be
expanded by the Coq user. Therefore, we know in advance all the Gallina tokens that can appear in a development, and we can
assign a concrete value to each of them. The function $[.]_{Gallina}: Gallina~tokens  \rightarrow \mathbb{Q}^-$ is an injective
function carefully defined to assign close values to similar Gallina tokens
and more distant numbers to unrelated tokens
--- see Appendix~\ref{sec:gallinasyntax} for the exact encoding.

The functions $[.]_{term}$ and $[.]_{type}$ are dynamically re-defined for every
library and every given proof stage, to adapt to the changing syntax.
In practice, there will be
new $[.]_{term}$ and $[.]_{type}$ functions computed whenever ML4PG is called.
This
brings the element of the ``acquired knowledge/experience'' to the machine-learning cycle, as will be formalised in the next section.
